\hydrogen{} and \carbon{} Network Strategies for the European Energy System

### 2024-08-26 report/submissions/journal/paper/paper.tex
adjust text and plots until new subsection

All scenarios share key energy system characteristics but differ in system costs and technology use (Fig.~\ref{fig:cost_bar}). Costs in the \baselinescenario{} are the highest at 779~bn€/a. Scenarios with carbon and hydrogen networks are more cost-effective, reducing costs by 3.1\% in the \carbongrid{}, 4.8\% in the \hydrogengrid{}, and 5.2\% in the \hybridscenario{}.

@@

System cost differences between the scenarios are driven by wind and solar, as well as carbon capture technologies. In scenarios with carbon and hydrogen networks, the model primarily reduces the reliance on high-cost DAC, from 392~Mt/a carbon removal in the \baselinescenario{} to 112~Mt/a in the \hybridscenario{} (see Fig.~\ref{fig:balance_captured_carbon}). At the same time, carbon capture from biomass combustion, gas-based industry processes and biogas-to-gas upgrading increase, the latter lowering the need for synthetic methane production.
% Another knock-on effect of less reliance on DAC is its reduced endogenous demand for dedicated renewables as well as heat pumps and biomass CHPs to supply power and heat.

@@

In the \carbonscenario, there are three main transport purposes of the low-cost carbon captured in Central and Eastern Europe (see Fig.~\ref{fig:balance_map}, top left). A large part of the carbon is transported to the Iberian Peninsula (8~kt/h), the British Isles (7.5~kt/h) and Denmark (2.8~kt/h), where it is used along with low-cost electrolytic hydrogen to produce Fischer-Tropsch fuels and methanol. A second part is transported to methanation plants in Italy and Greece (2.8~kt/h), where the synthetic methane is fed into the gas grid near large, geographically defined terminals where fossil gas is imported from outside Europe. Finally, a third part of captured carbon is transported directly to sites in the North Sea, Baltic Sea and Mediterranean Sea where it is sequestered. Almost none of the captured carbon is used locally at the point source. In addition to electrolysis in regions with good renewable resources to supply CU ($\sim$125~GW$_\text{el}$), the model places smaller amounts of electrolyzers where hydrogen is needed for industrial processes across Central Europe ($\sim$8.5~GW$_\text{el}$) (see bottom left). Hydrogen prices vary widely across the continent, with low price ``valleys'' at around 60~€/MWh, where the CU is located, and high price ``peaks'' in Central Europe at around 95~€/MWh. Despite the transport system, carbon prices still vary from 110~€/t in Central Europe to 160~€/t at sequestration sites.

In the \hydrogenscenario{}, the main hydrogen transportation routes go from Spain (50~GW) and the United Kingdom (18~GW) to Central Europe to supply local carbonaceous fuel production with captured carbon from point sources (see Fig~\ref{fig:balance_map}, bottom right). Also, the hydrogen network supplies spatially fixed hydrogen demand for industry in Central Europe (2.5~GW in total) at much lower costs than local electrolysis or SMR. Carbon sequestration is primarily located in Portugal where large DAC facilities operate (14~kt/h, see Fig~\ref{fig:balance_map}, top right). Carbon prices are low in Central Europe (110-120~€/t) and highest in sequestration regions (160~€/t). Hydrogen prices have flattened out at a relatively low level of €60~/MWh in all regions.

In both scenarios, CU is placed where material inputs, \carbon{} and \hydrogen{}, are provided at minimal cost, i.e., in price valleys of hydrogen and carbon in the \carbongrid{} and \hydrogenscenario{} respectively. In the \carbonscenario{}, CU co-locates with low-cost hydrogen production and uses transported carbon from the inland. In the \hydrogenscenario{}, CU co-locates with carbon point sources and uses transported low-cost hydrogen.
Thus, both scenarios effectively combine low-cost renewable hydrogen production and carbon management for combustion and spatially-fixed industrial processes.

However, there are individual cost-benefits. On the one hand, the carbon network enables the transport of low-cost captured carbon to sequestration sites, reducing the dependence on high-cost DAC near the coast which is present in the \hydrogenscenario{}. On the other hand, the \hydrogengrid{} is able to supply the spatially-fixed hydrogen demands in Central Europe with low-cost hydrogen imports, while the model in the \carbonscenario{} is forced to deploy electrolysis in regions with poor renewable resources as well as additional gas infrastructure in Central Europe. The latter aspect finally leads to lower system costs in the \hydrogenscenario{} than in the \carbonscenario{}.

@@

The \hybridscenario{} combines the advantages of hydrogen and carbon networks, resulting in the highest system cost reduction of 5.2\% (41~bn€/a) compared to the \baselinescenario{}. Again, the model transports low-cost hydrogen to the point sources in Central Europe to supply CU, but also transports low-cost carbon from point sources close to the shore to sequestration sites (see Fig.~\ref{fig:balance_map_full}). This provides low-cost hydrogen across the regions and reduces reliance on high-cost DAC.

For the carbon network, the model employs short and direct pipeline routes to transport carbon from inland point sources to nearby sequestration sites, for example in Northern Germany, Italy and Greece (see Fig.~\ref{fig:balance_map_full}, left). Additionally, the carbon network balances out small local variations in carbon demand and supply to produce CU, for example in the center of France.
The hydrogen routes are similar to those in the \hydrogenscenario{} with transports from Western regions (the Iberian Peninsula and the British Isles) to distributed CU production sites across the continent (see Fig.~\ref{fig:balance_map_full}, right).
While the topology of the hydrogen network (8~bn€/a investments) roughly corresponds to that of the \hydrogenscenario{} and is primarily driven by CU, the carbon network plays more of a complementary role and is primarily driven by CS (3~bn€/a investments, compare Fig.~\ref{fig:cost_bar_transmission}).

@@

In this combination, the networks enable all technologies that run on a high average capacity factor, i.e., above 70\%, to capture carbon to the full extent (see Fig.~\ref{fig:captureshare_line}). Except for SMR and gas-fired CHP which operate a few weeks annually, all emissions from carbon-emitting technologies are captured.
Thus, only a few DAC facilities are installed, such as those in southern Spain for methanolization plants. In comparison to the other scenarios, this leads to reduced power consumption and, therefore, less required wind and solar power deployment.

\subsection*{Higher sequestration limits reintroduce fossil fuels}

>>>>

All scenarios share key energy system characteristics but differ in system costs and technology use (Fig.~\ref{fig:cost_bar}). Costs in the \baselinescenario{} are the highest at 764~bn€/a. Scenarios with carbon and hydrogen networks are more cost-effective, reducing costs by 3.1\% in the \carbongrid{}, 4.7\% in the \hydrogengrid{}, and 5.3\% in the \hybridscenario{}.

@@

System cost differences between the scenarios are driven by wind and solar, as well as carbon capture technologies. In scenarios with carbon and hydrogen networks, the model primarily reduces the reliance on high-cost DAC, from 365~Mt/a carbon removal in the \baselinescenario{} to 113~Mt/a in the \hybridscenario{} (see Fig.~\ref{fig:balance_captured_carbon}). At the same time, carbon capture from biomass combustion, gas-based industry processes and biogas-to-gas upgrading increase, the latter lowering the need for synthetic methane production.
% Another knock-on effect of less reliance on DAC is its reduced endogenous demand for dedicated renewables as well as heat pumps and biomass CHPs to supply power and heat.

@@

In the \carbonscenario, there are three main transport purposes of the low-cost carbon captured in Central and Eastern Europe (see Fig.~\ref{fig:balance_map}, top left). A large part of the carbon is transported to the Iberian Peninsula (9.8~kt/h) and the British Isles (11.8~kt/h), where it is used along with low-cost electrolytic hydrogen to produce Fischer-Tropsch fuels and methanol. In Italy and Greece, a portion of the carbon (3.3~kt/h) is directed to methanation facilities where the synthetic methane is fed into the gas grid at key terminals that also receive fossil gas imports. Finally, a third part of captured carbon is transported directly to sites in the North Sea, Baltic Sea and Mediterranean Sea where it is sequestered. Almost none of the captured carbon is used locally at the point source. In addition to electrolysis in regions with good renewable resources to supply CU ($\sim$125~GW$_\text{el}$), the model places smaller amounts of electrolyzers where hydrogen is needed for industrial processes across Central Europe ($\sim$8~GW$_\text{el}$) (see bottom left). Hydrogen prices vary widely across the continent, with low price ``valleys'' at around 60~€/MWh, where the CU is located, and high price ``peaks'' in Central Europe at around 95~€/MWh. Despite the transport system, carbon prices still vary from 105~€/t in Central Europe to 160~€/t at sequestration sites.

In the \hydrogenscenario{}, the main hydrogen transportation route goes from Spain (66~GW) to Central Europe to supply local carbonaceous fuel production with captured carbon from point sources (see Fig~\ref{fig:balance_map}, bottom right). Also, the hydrogen network supplies spatially fixed hydrogen demand for industry in Central Europe (2.5~GW in total) at much lower costs than local electrolysis or SMR. Carbon sequestration takes place at several coastal sites with local DAC facilities. The largest carbon sink is located in Portugal (3.5~kt/h, see Fig~\ref{fig:balance_map}, top right). Carbon prices are low in Central Europe (110-120~€/t) and highest in sequestration regions (160~€/t). Hydrogen prices have flattened out at a relatively low level of €60~/MWh in all regions.

In both scenarios, CU is placed where material inputs, \carbon{} and \hydrogen{}, are provided at minimal cost, i.e., in price valleys of hydrogen and carbon in the \carbongrid{} and \hydrogenscenario{} respectively. In the \carbonscenario{}, CU co-locates with low-cost hydrogen production and uses transported carbon from the inland. In the \hydrogenscenario{}, CU co-locates with carbon point sources and uses transported low-cost hydrogen.
% Both scenarios effectively combine low-cost renewable hydrogen production and carbon management for combustion and spatially-fixed industrial processes.

We conclude, there are individual cost-benefits of the transport networks. On the one hand, the carbon network enables the transport of low-cost captured carbon to sequestration sites, reducing the dependence on high-cost DAC near the coast which is present in the \hydrogenscenario{}. On the other hand, the \hydrogengrid{} is able to supply the spatially-fixed hydrogen demands in Central Europe with low-cost hydrogen imports, while the model in the \carbonscenario{} is forced to deploy electrolysis in regions with poor renewable resources as well as additional gas infrastructure in Central Europe. The latter aspect finally leads to lower system costs in the \hydrogenscenario{} than in the \carbonscenario{}.

@@

The \hybridscenario{} combines the advantages of hydrogen and carbon networks, resulting in the highest system cost reduction of 5.3\% (41~bn€/a) compared to the \baselinescenario{}. Again, the model transports low-cost hydrogen to the point sources in Central Europe to supply CU, but also transports low-cost carbon from point sources close to the shore to sequestration sites (see Fig.~\ref{fig:balance_map_full}). This provides low-cost hydrogen across the regions and reduces reliance on high-cost DAC.

For the carbon network, the model employs short and direct pipeline routes to transport carbon from inland point sources to nearby sequestration sites, for example in Northern Germany, Italy and Greece (see Fig.~\ref{fig:balance_map_full}, left). In addition, pipelines transport rather small amounts of carbon from spatially fixed point sources to CU facilities. A notable outlier is the route from the Netherlands via the UK and France to large methanolisation and Fischer-Tropsch plants in northern Spain.
The hydrogen routes are similar to those in the \hydrogenscenario{} with transports from Western regions (the Iberian Peninsula and the British Isles) to distributed CU production sites across the continent (see Fig.~\ref{fig:balance_map_full}, right).
While the topology of the hydrogen network (1.2~bn€/a) roughly corresponds to that of the \hydrogenscenario{} and is primarily driven by CU, the carbon network (0.6~bn€/a) plays a secondary role and is driven by CS (compare Fig.~\ref{fig:cost_bar_transmission}).

@@

In this setup, the networks enable all carbon-emitting technologies that operate at a high average capacity factor, i.e., above 70\%, to capture most of their carbon (see Fig.~\ref{fig:captureshare_line}). Except for SMR and gas-fired CHP which operate a few weeks annually, more than 90\% emissions are captured.
Due to the better integration of point sources, only a few DAC facilities are installed, such as those used for methanolisation in southern Spain. This, in turn, leads to reduced power consumption and less required wind and solar power deployment than in the other scenarios.

\subsection*{Higher sequestration limits reintroduce fossil fuels}
\label{sec:higher_sequestration}
Carbon management and hard-to-abate sectors of the optimized system are strongly driven by the model's limited sequestration potential of 200~Mt/a. In the Hybrid scenario, this limitations imposes a shadow-price of 225€/t to the system, meaning the system would gain 225€ from an additional tonne of sequestration potential. As soon as the sequestration limit is increased, the model starts to reintroduce fossil fuels as a primary energy source in the the aviation and industrial sectors.
